---
title: "Adegenet From VCF"
output:
  html_document:
    df_print: paged
---

## Load libraries and set working directory
```{r load-libs, message=FALSE, warning=FALSE}
library(adegenet)
library(vcfR)
library(pegas)
library(hierfstat)
library(ggplot2)
library(viridis)
library(tidyverse)     # for later chunks
library(adegraphics)

knitr::opts_chunk$set(fig.path = "plots_LD/", dev = c("png", "svg"), dpi = 500, cache = TRUE)
```

## Import data, make `genind`, and attach populations
```{r import-data, message=FALSE, warning=FALSE}
# ---- paths ----
vcfFile     <- "/scratch/Raccoon_andrea/gabe_trial/data/vcf/Raccoon.filtered.vcf"
labels_path <- "/scratch/Raccoon_andrea/gabe_trial/scripts/PopulationGenetics/labels.csv"

# ---- VCF -> loci -> genind ----
allPop.vcf    <- read.vcf(vcfFile, to = 38500) #39000 since the 
allPop.genind <- loci2genind(allPop.vcf)

# ---- read labels (Sample, Location) and clean ----
lab <- read.csv(labels_path, stringsAsFactors = FALSE)
lab$Sample   <- trimws(as.character(lab$Sample))
lab$Location <- trimws(as.character(lab$Location))
lab$Location[is.na(lab$Location) | lab$Location == ""] <- "Unknown"

# ---- align label rows to genind individual order ----
ids <- adegenet::indNames(allPop.genind)
ids <- trimws(ids)

idx <- match(ids, lab$Sample)
if (any(is.na(idx))) {
  miss <- ids[is.na(idx)]
  stop(sprintf("These VCF/genind sample IDs are missing from labels.csv: %s",
               paste(miss, collapse = ", ")))
}
lab_ord <- lab[idx, , drop = FALSE]  # same order as genind
stopifnot(identical(ids, lab_ord$Sample))

# ---- attach population factor ----
adegenet::pop(allPop.genind) <- factor(lab_ord$Location)

# quick sanity prints
# print(table(pop(allPop.genind)))
# print(nInd(allPop.genind))
```


## Start analysis

### Bartlett test of homogeneity of variances
```{r bartlett-test, cache=TRUE}
allPop.summary <- summary(allPop.genind)
bartlett.test(list(allPop.summary$Hexp, allPop.summary$Hobs))
t.test(allPop.summary$Hexp, allPop.summary$Hobs,
       paired = TRUE, var.equal = TRUE, alternative = "greater")
```

### F-stats

#### Test of Hardy–Weinberg Equilibrium
```{r hwt, message=FALSE, warning=FALSE, cache=TRUE}
allPop.hwt <- hw.test(allPop.genind, B = 0)
colMeans(allPop.hwt)
```

#### General Fst & Fis
```{r fstats, cache=TRUE}
allPop.fstats <- wc(allPop.genind)
print(paste("FIS:", allPop.fstats$FIS))
print(paste("FST:", allPop.fstats$FST))

allPop.basics <- basic.stats(allPop.genind)
allPop.basics$overall
```

#### Per-population diversity (Ho, He, FIS)
```{r per-pop-diversity, cache=TRUE}
# Calculate basic stats per population (matrices: loci x populations)
pop_stats <- hierfstat::basic.stats(allPop.genind)

# Per-population means across loci
pop_ho  <- colMeans(pop_stats$Ho, na.rm = TRUE)   # observed het
pop_he  <- colMeans(pop_stats$Hs, na.rm = TRUE)   # expected het
pop_fis <- colMeans(pop_stats$Fis, na.rm = TRUE)  # inbreeding coeff

# Fallback for FIS if needed (e.g., divisions produce NaN for some loci/pops)
if (length(pop_fis) != length(pop_ho) || any(is.na(pop_fis))) {
  pop_fis_alt <- colMeans(1 - (pop_stats$Ho / pop_stats$Hs), na.rm = TRUE)
  # use alt where primary is NA
  nm <- names(pop_fis)
  pop_fis[is.na(pop_fis)] <- pop_fis_alt[is.na(pop_fis)]
  names(pop_fis) <- nm
}

# Ensure consistent ordering of vectors
Population <- names(pop_ho)

# Optional: sample sizes per population
n_by_pop <- as.data.frame(table(Population = as.character(pop(allPop.genind))))
names(n_by_pop)[2] <- "N"

# Combine to a tidy table
pop_div_df <- tibble::tibble(
  Population = Population,
  Ho  = round(pop_ho[Population], 3),
  He  = round(pop_he[Population], 3),
  FIS = round(pop_fis[Population], 3)
) %>%
  dplyr::left_join(n_by_pop, by = "Population") %>%
  dplyr::arrange(Population)

# Write to scratch directory and print
readr::write_csv(pop_div_df, "/scratch/Raccoon_andrea/gabe_trial/results/PerPopulation_Diversity.csv")
pop_div_df

# Quick visual check (optional)
ggplot(pop_div_df, aes(He, Ho, label = Population)) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(nudge_y = 0.02, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Expected heterozygosity (He)",
       y = "Observed heterozygosity (Ho)",
       title = "Per-population heterozygosity across counties") +
  theme_minimal(base_size = 12)
```


#### Bootstraps
```{r bootstraps, cache=TRUE}
betas(allPop.genind, nboot = 102, lim = c(0.025, 0.975), diploid = TRUE, betaijT = FALSE)
```

### Inbreeding
```{r inbreeding, message=FALSE, warning=FALSE, cache=TRUE}
allPop.inbr <- inbreeding(allPop.genind, N = 100)
allPop.FIS  <- sapply(allPop.inbr, mean)

hist(allPop.FIS, col = "orange", xlab = "mean value of F",
     main = "Distribution of mean F across individuals")
```


#Helper (put once, before PCA/PCoA)
```{r plot-helper, include=FALSE}
# -------------------------------------------------------------------------
# Helper (put once, before PCA/PCoA)
# -------------------------------------------------------------------------
render_sclass <- function(dfxy, fac, main, xlab, ylab) {
  # 1) Try adegraphics
  g1 <- try(adegraphics::s.class(dfxy = dfxy, fac = fac, xax = 1, yax = 2,
                                 psub.text = main, xlab = xlab, ylab = ylab),
            silent = TRUE)
  if (!inherits(g1, "try-error")) return(invisible(NULL))

  # 2) Fallback: ade4
  g2 <- try({
    ade4::s.class(dfxy, fac = fac, xax = 1, yax = 2,
                  col = adegenet::funky(nlevels(fac)),
                  cpoint = 0.7, clabel = 0)
    title(main = main, xlab = xlab, ylab = ylab)
  }, silent = TRUE)
  if (!inherits(g2, "try-error")) return(invisible(NULL))

  # 3) Fallback: ggplot
  df <- data.frame(A1 = dfxy[,1], A2 = dfxy[,2], Pop = fac)
  print(
    ggplot2::ggplot(df, ggplot2::aes(A1, A2, color = Pop)) +
      ggplot2::stat_ellipse(type = "norm", linewidth = 0.5) +
      ggplot2::geom_point(size = 0.7) +
      ggplot2::labs(title = main, x = xlab, y = ylab) +
      ggplot2::theme_minimal()
  )
}

allPop.pcoa <- dudi.pco(dist(allPop.genind), scannf = FALSE, nf = 15)
var.explained.pcoa <- round(allPop.pcoa$eig / sum(allPop.pcoa$eig) * 100, 1)

dfxy <- as.data.frame(allPop.pcoa$li[, 1:2, drop = FALSE])
names(dfxy) <- c("Axis1", "Axis2")
fac  <- droplevels(as.factor(pop(allPop.genind)))

render_sclass(
  dfxy, fac,
  main = "PCoA — colored by County",
  xlab = paste0("Axis 1 (", var.explained.pcoa[1], "%)"),
  ylab = paste0("Axis 2 (", var.explained.pcoa[2], "%)")
)
```

```{r region_map, message=FALSE}
# (your region_map code)
region_map <- tibble::tribble(
  ~Population,   ~Region,
  # Southwest
  "Baldwin",     "SW",
  "Mobile",      "SW",
  "Monroe",      "SW",
  "Marengo",     "SW",
  "Dallas",      "SW",

  # Southeast
  "Houston",     "SE",
  "Dale",        "SE",
  "Coffee",      "SE",
  "Geneva",      "SE",
  "Pike",        "SE",
  "Barbour",     "SE",

  # Black Belt / Central
  "Montgomery",  "BB",
  "Elmore",      "BB",
  "Autauga",     "BB",
  "Lee",         "BB",
  "Chilton",     "BB",
  "Macon",       "BB",

  # Northeast
  "Calhoun",     "NE",
  "Etowah",      "NE",
  "Cherokee",    "NE",
  "DeKalb",      "NE",
  "Blount",      "NE",
  "Marshall",    "NE",
  "StClair",     "NE",
  "Talladega",   "NE",
  "Shelby",      "NE",
  "Jefferson",   "NE",
  "Cullman",     "NE",

  # Northwest (no Fayette, Tuscaloosa)
  "Lauderdale",  "NW",
  "Colbert",     "NW",
  "Lawrence",    "NW",
  "Madison",     "NW",
  "Morgan",      "NW",
  "Jackson",     "NW"
)

region_cols <- c(
  SW="#1b9e77", BB="#d95f02", NE="#7570b3", SE="#e7298a", NW="#66a61e", Other="#a6761d"
)

# Check for unmapped counties
mapped <- unique(region_map$Population)
unmapped <- setdiff(unique(as.character(pop(allPop.genind))), mapped)
if (length(unmapped)) message("Unmapped counties (will be 'Other' in plots): ",
                              paste(sort(unmapped), collapse=", "))


po <- allPop.pcoa$li
pcoa_df <- tibble::tibble(
  Sample = rownames(po),
  Axis1 = as.numeric(po[,1]),
  Axis2 = as.numeric(po[,2]),
  Pop = as.character(pop(allPop.genind))
)
```

```{r pcoa_county, message=FALSE}
# (your PCoA-by-county plot)

ggplot(pcoa_df, aes(Axis1, Axis2, color = Pop)) +
  stat_ellipse(type = "norm", linewidth = 0.4, alpha = 0.6) +
  geom_point(size = 1) +
  labs(
    title = "PCoA — colored by County (Population)",
    x = paste0("Axis 1 (", round(allPop.pcoa$eig[1] / sum(allPop.pcoa$eig) * 100, 1), "%)"),
    y = paste0("Axis 2 (", round(allPop.pcoa$eig[2] / sum(allPop.pcoa$eig) * 100, 1), "%)"),
    color = "County"
  ) +
  theme_minimal(base_size = 11) +
  guides(color = guide_legend(ncol = 2, override.aes = list(size = 3)))

pcoa_reg <- pcoa_df %>%
  dplyr::left_join(region_map, by = c("Pop" = "Population")) %>%
  dplyr::mutate(Region = if_else(is.na(Region), "Other", Region),
                Region = factor(Region, levels = c("SW","SE","BB","NE","NW","Other")))
```

```{r pcoa_region, message=FALSE}
# (your PCoA-by-region plot)

ggplot(pcoa_reg, aes(Axis1, Axis2, color = Region)) +
  stat_ellipse(type = "norm", linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1.2) +
  scale_color_manual(values = region_cols, name = "Region", drop = FALSE) +
  labs(
    title = "PCoA — colored by Region",
    x = paste0("Axis 1 (", round(allPop.pcoa$eig[1] / sum(allPop.pcoa$eig) * 100, 1), "%)"),
    y = paste0("Axis 2 (", round(allPop.pcoa$eig[2] / sum(allPop.pcoa$eig) * 100, 1), "%)")
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")


out_dir <- "/scratch/Raccoon_andrea/gabe_trial/results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
cat("N individuals:", nInd(allPop.genind), "  N loci:", nLoc(allPop.genind), "\n")
```




### H1 — Pairwise FST (all counties) + your target pairs

```{r H1-pairwise-fst}
dat_hier <- hierfstat::genind2hierfstat(allPop.genind)
pw_wc    <- hierfstat::pairwise.WCfst(dat_hier)  # matrix Pop x Pop

# All pairs in long form
pw_tbl <- as.data.frame(as.table(as.matrix(pw_wc)), stringsAsFactors = FALSE)
names(pw_tbl) <- c("PopA","PopB","Fst")
pw_tbl <- pw_tbl |>
  dplyr::filter(PopA != PopB, !is.na(Fst)) |>
  dplyr::mutate(PopA = as.character(PopA), PopB = as.character(PopB)) |>
  dplyr::filter(PopA < PopB) |>
  dplyr::arrange(dplyr::desc(Fst))
readr::write_csv(pw_tbl, file.path(out_dir, "H1_pairwise_FST_all_pairs.csv"))
head(pw_tbl, 12)
```

```{r H1-pairwise-summary, message=FALSE}
# Summarize pairwise FST among Alabama counties
upper_vals <- pw_wc[upper.tri(pw_wc, diag = FALSE)]
pairwise_summary <- tibble::tibble(
  median = median(upper_vals, na.rm = TRUE),
  min    = min(upper_vals, na.rm = TRUE),
  max    = max(upper_vals, na.rm = TRUE),
  q25    = quantile(upper_vals, 0.25, na.rm = TRUE, names = FALSE),
  q75    = quantile(upper_vals, 0.75, na.rm = TRUE, names = FALSE),
  n_pairs = sum(upper.tri(pw_wc, diag = FALSE))
)

readr::write_csv(pairwise_summary, file.path(out_dir, "H1_pairwise_FST_summary.csv"))
pairwise_summary
```

```{r H1-target-pairs, message=FALSE}
pairs_h1 <- tibble::tribble(
  ~A,           ~B,           ~barrier,
  "Baldwin",    "Mobile",     TRUE,   # river barrier
  "Montgomery", "Elmore",     FALSE,  # same side, control
  "Autauga",    "Chilton",    FALSE,  # same side, control
  "Autauga",    "Montgomery", TRUE,   # across Alabama River
  "Chilton",    "Elmore",     TRUE,   # across Coosa
  "Cherokee",   "Calhoun",    TRUE,
  "DeKalb",     "Calhoun",    TRUE,
  "Etowah",     "Calhoun",    TRUE,
  "DeKalb",     "Etowah",     FALSE,  # control
  "Talladega",  "Shelby",     TRUE    # across Coosa
)

# normalize order to avoid A/B vs B/A duplicates
pairs_h1 <- pairs_h1 |>
  dplyr::rowwise() |>
  dplyr::mutate(
    A_norm = min(A, B),
    B_norm = max(A, B)
  ) |>
  dplyr::ungroup() |>
  dplyr::distinct(A_norm, B_norm, .keep_all = TRUE) |>
  dplyr::select(A = A_norm, B = B_norm, barrier)

all_pops <- intersect(rownames(pw_wc), colnames(pw_wc))
missing  <- setdiff(unique(c(pairs_h1$A, pairs_h1$B)), all_pops)
if (length(missing)) warning("Missing populations in FST matrix: ", paste(missing, collapse=", "))

pairs_tbl <- pairs_h1 |>
  dplyr::filter(A %in% all_pops, B %in% all_pops) |>
  dplyr::rowwise() |>
  dplyr::mutate(Fst = as.numeric(pw_wc[A, B])) |>
  dplyr::ungroup()

readr::write_csv(pairs_tbl, file.path(out_dir, "H1_pairwise_FST_target_pairs.csv"))
pairs_tbl
```

```{r H1-target-pairs-ci, message=FALSE}
pair_cis <- vector("list", nrow(pairs_tbl))
nboot_pair <- 200   # use 1000+ for final (speed vs stability)

for (i in seq_len(nrow(pairs_tbl))) {
  A <- pairs_tbl$A[i]; B <- pairs_tbl$B[i]
  sub <- allPop.genind[pop(allPop.genind) %in% c(A,B)]
  dh  <- hierfstat::genind2hierfstat(sub)
  loc_cols <- 2:ncol(dh)
  reps <- numeric(nboot_pair)
  for (b in seq_len(nboot_pair)) {
    samp <- sample(loc_cols, length(loc_cols), replace = TRUE)
    reps[b] <- suppressWarnings(as.numeric(hierfstat::wc(dh[, c(1, samp)])$FST))
  }
  ci <- stats::quantile(reps, c(.025,.975), na.rm=TRUE, names=FALSE)
  pair_cis[[i]] <- tibble::tibble(A=A, B=B, Fst_CI_lower=ci[1], Fst_CI_upper=ci[2])
}
pair_ci_tbl <- dplyr::bind_rows(pair_cis)

pairs_tbl <- dplyr::left_join(pairs_tbl, pair_ci_tbl, by=c("A","B"))
readr::write_csv(pairs_tbl, file.path(out_dir, "H1_pairwise_FST_target_pairs_withCI.csv"))
pairs_tbl
```

```{r H1-barrier-test, message=FALSE}
# One-sided Wilcoxon: barrier > control
h1_wilcox <- wilcox.test(Fst ~ barrier, data = pairs_tbl, alternative = "greater", exact = FALSE)

# Bootstrap CI for mean difference (barrier - control)
set.seed(123)
B <- 5000
diffs <- numeric(B)
xb <- pairs_tbl$Fst[pairs_tbl$barrier]
xc <- pairs_tbl$Fst[!pairs_tbl$barrier]
for (b in seq_len(B)) {
  diffs[b] <- mean(sample(xb, length(xb), replace = TRUE)) -
              mean(sample(xc, length(xc), replace = TRUE))
}
h1_boot_ci <- stats::quantile(diffs, c(.025, .975), names = FALSE)

h1_results <- tibble::tibble(
  test = "Wilcoxon (barrier > control)",
  p_value = h1_wilcox$p.value,
  mean_diff = mean(xb) - mean(xc),
  diff_CI_lower = h1_boot_ci[1],
  diff_CI_upper = h1_boot_ci[2]
)
readr::write_csv(h1_results, file.path(out_dir, "H1_barrier_effect_test.csv"))
h1_results
```


#### Quick H1 figure (barriers vs non-barriers)

```{r H1-barplot, fig.height=4, fig.width=7}
ggplot2::ggplot(pairs_tbl,
                ggplot2::aes(x = reorder(paste(A, B, sep=" vs "), Fst),
                             y = Fst, fill = barrier)) +
  ggplot2::geom_col() +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = c(`TRUE`="#cd3e3e", `FALSE`="#3e7ccd"),
                             name = "River barrier") +
  ggplot2::labs(x = NULL, y = "Weir–Cockerham FST",
                title = "Target county pairs") +
  ggplot2::theme_minimal(base_size = 12)
```


## H3 Region AMOVA — updated regions
## (uses out_dir already defined earlier)
```{r region_mappp, fig.height=4, fig.width=7}
# ----- Region map (updated) -----
region_map <- tibble::tribble(
  ~Population,  ~Region,
  # Southwest (Mobile basin / Coastal Plain)
  "Baldwin",    "SW",
  "Mobile",     "SW",
  "Monroe",     "SW",
  "Marengo",    "SW",
  "Dallas",     "SW",

  # Southeast (SE Coastal Plain)
  "Houston",    "SE",
  "Dale",       "SE",
  "Coffee",     "SE",
  "Geneva",     "SE",
  "Pike",       "SE",
  "Barbour",    "SE",

  # Black Belt / Central
  "Montgomery", "BB",
  "Elmore",     "BB",
  "Autauga",    "BB",
  "Lee",        "BB",
  "Chilton",    "BB",
  "Macon",      "BB",

  # Northeast (Appalachian foothills / Ridge–Valley)
  "Calhoun",    "NE",
  "Etowah",     "NE",
  "Cherokee",   "NE",
  "DeKalb",     "NE",
  "Blount",     "NE",
  "Marshall",   "NE",
  "StClair",    "NE",
  "Talladega",  "NE",
  "Shelby",     "NE",
  "Jefferson",  "NE",
  "Cullman",    "NE",

  # Northwest (Tennessee Valley / Highland Rim)
  "Lauderdale", "NW",
  "Colbert",    "NW",
  "Lawrence",   "NW",
  "Madison",    "NW",
  "Morgan",     "NW",
  "Jackson",    "NW"
)

# ----- Build meta and join regions (with a couple of name fixes) -----
pop_names <- as.character(adegenet::pop(allPop.genind)) |> trimws()
pop_names <- dplyr::recode(pop_names,
                           "St. Clair" = "StClair",
                           "Montomery" = "Montgomery")

meta <- tibble::tibble(Sample = adegenet::indNames(allPop.genind),
                       Population = pop_names) |>
        dplyr::left_join(region_map, by = "Population")

# Warn about anything unmapped
unmapped <- sort(unique(meta$Population[is.na(meta$Region)]))
if (length(unmapped)) {
  message("Unmapped counties (dropped from H3 AMOVA): ",
          paste(unmapped, collapse = ", "))
}

# Keep only samples with a defined Region
keep <- !is.na(meta$Region)
g2   <- allPop.genind[keep]
meta <- meta[keep, , drop = FALSE]

# Attach hierarchical strata and set population level
adegenet::strata(g2) <- data.frame(
  Region     = factor(meta$Region, levels = c("SW","SE","BB","NE","NW")),
  Population = meta$Population
)
adegenet::setPop(g2) <- ~Population

# Quick sanity check: counts per Region
print(dplyr::count(meta, Region, sort = TRUE))

# ----- AMOVA (999 perms) -----
a2 <- poppr::poppr.amova(g2, ~Region/Population, nperm = 999)

# Tidy variance components (% of total), robust to column naming
comp_df <- as.data.frame(a2$componentsofcovariance)
comp_df$Level <- rownames(comp_df)
var_col <- dplyr::case_when(
  "Variance" %in% names(comp_df) ~ "Variance",
  "Sigma"    %in% names(comp_df) ~ "Sigma",
  TRUE ~ names(comp_df)[which(sapply(comp_df, is.numeric))[1]]
)

tab <- tibble::tibble(
  Level    = comp_df$Level,
  Variance = as.numeric(comp_df[[var_col]])
) |>
  dplyr::mutate(Percent = 100 * Variance / sum(Variance))

readr::write_csv(tab, file.path(out_dir, "H3_AMOVA_hierarchical.csv"))

# Save Φ-statistics / permutation p-values if available
if (!is.null(a2$statphi)) {
  phi_tab <- as.data.frame(a2$statphi)
  phi_tab$Level <- rownames(phi_tab)
  readr::write_csv(phi_tab, file.path(out_dir, "H3_AMOVA_phi_pvals.csv"))
}

tab
```



