cat > Hypotheses-analysis.Rmd <<'EOF'
---
title: "Raccoon genetics: H1–H3 analyses"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(adegenet)
  library(pegas)
  library(hierfstat)
  library(poppr)
  library(tidyverse)   # dplyr, tidyr, readr, stringr, tibble, ggplot2
  library(vegan)       # for Mantel (H3, optional)
})

set.seed(42)
knitr::opts_chunk$set(fig.path = "plots_LD/", dev = c("png", "svg"), dpi = 300, cache = FALSE)

# Paths (EDIT if needed)
vcfFile     <- "/scratch/Raccoon_andrea/gabe_trial/data/vcf/Raccoon.filtered.vcf"
labels_path <- "/scratch/Raccoon_andrea/gabe_trial/scripts/PopulationGenetics/labels.csv"
out_dir     <- "/scratch/Raccoon_andrea/gabe_trial/results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Hypothesis 1 pairs
pairs_h1 <- tibble::tribble(
  ~A,           ~B,        ~barrier,
  "Baldwin",    "Mobile",  TRUE,
  "Montgomery", "Elmore",  TRUE,
  "Autauga",    "Chilton", TRUE,
  "Cherokee",   "Calhoun", TRUE,
  "DeKalb",     "Calhoun", FALSE
)
``` 

###Load VCF → genind and attach populations
```{r genind, message=FALSE, warning=FALSE}
message("Reading VCF (pegas::read.vcf)...")
# read only the first 300 loci (fast test)
allPop.vcf <- read.vcf(vcfFile, to = 300)
#allPop.vcf    <- read.vcf(vcfFile)   # add 'to=30000' if you want to limit loci
message("Converting to genind...")
allPop.genind <- loci2genind(allPop.vcf)

# Read & clean labels (Sample, Location)
lab <- read.csv(labels_path, stringsAsFactors = FALSE)
lab$Sample   <- stringr::str_squish(as.character(lab$Sample))
lab$Location <- stringr::str_squish(as.character(lab$Location))
lab$Location[is.na(lab$Location) | lab$Location == ""] <- "Unknown"
lab$Location <- dplyr::recode(lab$Location,
  "St Clair" = "StClair",
  "St. Clair" = "StClair",
  "De Kalb" = "DeKalb",
  "Dekalb"  = "DeKalb",
  .default = lab$Location
)
lab <- dplyr::distinct(lab, Sample, .keep_all = TRUE)

# Align labels to genind order and attach pop()
ids <- adegenet::indNames(allPop.genind) %>% stringr::str_squish()
idx <- match(ids, lab$Sample)
if (any(is.na(idx))) {
  stop("Missing samples in labels.csv: ", paste(ids[is.na(idx)], collapse=", "))
}
lab_ord <- lab[idx, , drop = FALSE]
stopifnot(identical(ids, lab_ord$Sample))
adegenet::pop(allPop.genind) <- factor(lab_ord$Location)

# quick sanity
data.frame(N_individuals = nInd(allPop.genind),
           N_pops = nPop(allPop.genind)) %>% knitr::kable()
```

###H2 — Per-county diversity (Ho/He/FIS)
```{r cdiversity, message=FALSE, warning=FALSE}
bs <- basic.stats(allPop.genind)
div_summary <- tibble(
  Population = rownames(bs$perloc),
  Ho  = rowMeans(bs$Ho, na.rm = TRUE),
  He  = rowMeans(bs$Hs, na.rm = TRUE),
  FIS = rowMeans(bs$Fis, na.rm = TRUE),
  n_ind = as.integer(table(pop(allPop.genind))[Population])
) %>% arrange(Population)

write.csv(div_summary, file.path(out_dir, "H2_diversity_summary_by_county.csv"), row.names = FALSE)
div_summary %>% dplyr::arrange(desc(He)) %>% head(10) %>% knitr::kable()
```


###H1 — Pairwise FST (all counties) + your target pairs
```{r H1-pairwise-fst, message=FALSE}
dat_hier <- hierfstat::genind2hierfstat(allPop.genind)

# All pairwise WC FST as a matrix
pw_wc <- hierfstat::pairwise.WCfst(dat_hier)

# Robust reshape to long format with explicit column names
pw_tbl <- as.data.frame(as.table(as.matrix(pw_wc)), stringsAsFactors = FALSE)
names(pw_tbl) <- c("PopA", "PopB", "Fst")

# Keep only unique pairs (upper triangle), drop self-pairs and NAs
pw_tbl <- pw_tbl %>%
  dplyr::filter(PopA != PopB, !is.na(Fst)) %>%
  dplyr::mutate(PopA = as.character(PopA), PopB = as.character(PopB)) %>%
  dplyr::filter(PopA < PopB) %>%              # unique direction
  dplyr::arrange(dplyr::desc(Fst))

# Save + quick preview
write.csv(pw_tbl, file.path(out_dir, "H1_pairwise_FST_all_pairs.csv"), row.names = FALSE)
pw_tbl %>% head(15) %>% knitr::kable()
```


###And update the **target pairs** extraction to guard against typos/missing pops:

```{r H1-target-pairs, message=FALSE}
# Check that all requested pops exist in the FST matrix
all_pops <- intersect(rownames(pw_wc), colnames(pw_wc))
missing <- setdiff(unique(c(pairs_h1$A, pairs_h1$B)), all_pops)
if (length(missing)) {
  warning("These populations are not in the FST matrix: ", paste(missing, collapse = ", "))
}

pairs_tbl <- pairs_h1 %>%
  dplyr::filter(A %in% all_pops, B %in% all_pops) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Fst = as.numeric(pw_wc[A, B])) %>%
  dplyr::ungroup()

write.csv(pairs_tbl, file.path(out_dir, "H1_pairwise_FST_target_pairs.csv"), row.names = FALSE)
pairs_tbl %>% knitr::kable()
```

###H1 — AMOVA per pair (Population = county)
```{r amova, message=FALSE, warning=FALSE}
amova_rows <- list()
for (i in seq_len(nrow(pairs_h1))) {
  A <- pairs_h1$A[i]; B <- pairs_h1$B[i]; br <- pairs_h1$barrier[i]
  sub <- allPop.genind[pop(allPop.genind) %in% c(A, B)]
  poppr::strata(sub) <- data.frame(Location = droplevels(pop(sub)))
  poppr::setPop(sub) <- ~Location
  a <- poppr::poppr.amova(sub, ~Location, nperm = 999)
  perc <- 100 * a$componentsofcovariance / sum(a$componentsofcovariance)
  amova_rows[[i]] <- tibble(
    Pair = paste(A, B, sep=" vs "),
    barrier = br,
    Among_Pops_sigma = a$componentsofcovariance[1],
    Among_Pops_pct   = perc[1],
    Within_sigma     = a$componentsofcovariance[2],
    Within_pct       = perc[2],
    P_value          = a$statphi$`Pr(>Phi)`[1]
  )
}
amova_tbl <- bind_rows(amova_rows)
write.csv(amova_tbl, file.path(out_dir, "H1_AMOVA_by_pair.csv"), row.names = FALSE)
amova_tbl %>% knitr::kable(digits = 4)
```
###H1 — Overall WC FST + 95% CI (bootstrap)
```{r bootstrap, message=FALSE, warning=FALSE}
wc_overall <- wc(allPop.genind)$FST
ci_overall <- boot.vc(dat = dat_hier[, -1], pop = dat_hier[, 1], nboot = 1000)$ci
overall_tbl <- tibble(Estimate_FST = wc_overall,
                      CI_lower = ci_overall[1], CI_upper = ci_overall[2])
write.csv(overall_tbl, file.path(out_dir, "H1_overall_FST_CI.csv"), row.names = FALSE)
overall_tbl %>% knitr::kable(digits = 5)
```

###H3 — Mantel (IBD) scaffold (optional; needs coordinates)
```{r mantel, message=FALSE, warning=FALSE}
coords_path <- file.path(dirname(labels_path), "sample_coords_template.csv")
if (!file.exists(coords_path)) {
  coords <- tibble::tibble(Sample = lab_ord$Sample, lon = NA_real_, lat = NA_real_)
  write.csv(coords, coords_path, row.names = FALSE)
  message("Wrote coordinate template: ", coords_path,
          "\nFill lon/lat (WGS84) to enable Mantel (IBD) analysis.")
} else {
  coords <- read.csv(coords_path)
  if (all(!is.na(coords$lon)) && all(!is.na(coords$lat))) {
    coords <- coords[match(ids, coords$Sample), c("lon","lat")]
    Dgen <- dist(allPop.genind)
    Dgeo <- dist(as.matrix(coords))
    mantel_res <- vegan::mantel(Dgen, Dgeo, permutations = 999)
    saveRDS(mantel_res, file.path(out_dir, "H3_mantel_result.rds"))
    mantel_res
  } else {
    message("Coords file exists but contains NA lon/lat — skipping Mantel.")
  }
}
```
